---
layout: default
tags: [regression, model-checking ]
---

```{r echo=FALSE}
library(lme4)
library(lattice)
#library(latticeExtra)
opts_chunk$set(dpi = 144)
opts_chunk$set(out.width = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
source("../../src/import-data.R")
source("../../src/fit-glms.R")
```

```{r}
gene.ids <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(gene.ids) <- gene.ids
```

Prepare data frame including all `r length(gene.ids)` selected genes and read formulas (linear predictors) for models like the relatively simple $$M1$$ and the more complex $$M3$$:

```{r cache=FALSE}
dat <- merge.data(gene.ids = gene.ids)
fm <- read.csv(file = "../../results/M-formulas.csv", stringsAsFactors = FALSE)
```

### Results under $$M3$$

Fit models; note messages warning problems with convergence (slow convergence) for *wnlm.Q* and *logi.S*

```{r cache=TRUE}
M3 <- list()
M3$unlm.Q <- lmer(reformulate(fm$M3[3], response = "Q"), data = dat)
M3$wnlm.Q <- lmer(reformulate(fm$M3[3], response = "Q"), data = dat, weights = N)
M3$unlm.S <- lmer(reformulate(fm$M3[3], response = "S"), data = dat)
M3$logi.S <- glmer(reformulate(fm$M3[3], response = "H.N"), data = dat, family = binomial)
```

```{r}
get.diagnostic.data <- function(lM, sel.col = 2:7) {
    helper <- function(x)
        cbind(data.frame(Residual = residuals(m <- lM[[x]]), Fitted.value = predict(m), Family = x),
              model.frame(m)[sel.col])
    l <- lapply(names(lM), helper)
    do.call(rbind, l)
}
```

#### Normality of residuals

The distribution of residuals under all four model families shows smaller or larger departure from normality.  *unlm.Q* is the closest to standard normal distribution and *wnlm.Q* is equally normal (or perhaps slightly less so).  The distribution under *logi.S* is strikingly far from normal.  So is it under *unlm.S*, for which additionally the scale of distribution is diminished (compare slope of black diagonal to the overall slope of the blue curve) indicating that the variance estimation---which is based on the residuals---is strongly biased downwards.

```{r qqplot-families-M3}
diag.M3 <- get.diagnostic.data(M3)
arg <- list(main = "Normality of residuals", xlab = "normal quantiles", ylab = "empirical quantiles", pch = "+")
qqmath(~ Residual | Family, data = diag.M3, scales = list(y = list(relation = "free")),
       xlab = arg$xlab, ylab = arg$ylab, main = arg$main, pch = arg$pch, abline = c(0, 1))
```

```{r qqplot-genes-unlm-Q-M3}
qqmath(~ Residual | Gene, data = diag.M3, subset = Family == "unlm.Q", xlab = arg$xlab, ylab = arg$ylab, main = paste(arg$main, ": unlm.Q"), pch = arg$pch, abline = c(0, 1))
```

#### Homogeneity of error

Inspecting the homogeneity of error variance (or equivalently standard deviation) leads to the same conclusion as above regarding the relative goodness of fit of the various model families:  *unlm.Q* and *wnlm.Q* are the best (the most homoscedastic) while *unlm.S* and *logi.S* show systematic relationships between error and fitted value indicating poorer fit.

```{r scedasticity-families-M3}
arg <- list(main = "Error variance", xlab = "fitted value", ylab = expression(sqrt(residual)), pch = "+")
xyplot(sqrt(abs(Residual)) ~ Fitted.value | Family, data = diag.M3, scales = list(relation = "free"),
       xlab = arg$xlab, ylab = arg$ylab, main = arg$main, pch = arg$pch)
```

```{r scedasticity-genes-unlm-Q-M3}
xyplot(sqrt(abs(Residual)) ~ Fitted.value | Gene, data = diag.M3, subset = Family == "unlm.Q", scales = list(x = list(relation = "free", draw = FALSE)), xlab = arg$xlab, ylab = arg$ylab, main = paste(arg$main, ": unlm.Q"), pch = arg$pch)
```

### Results under a simpler model, $$M1$$

Very similar patterns are seen under $$M1$$ to those under $$M3$$.  The fitting of the respective models from the *wnlm.Q* and *logi.S* families again converges with problems.

```{r cache=TRUE, echo=TRUE}
M1 <- list()
M1$unlm.Q <- lmer(reformulate(fm$M1[3], response = "Q"), data = dat)
M1$wnlm.Q <- lmer(reformulate(fm$M1[3], response = "Q"), data = dat, weights = N)
M1$unlm.S <- lmer(reformulate(fm$M1[3], response = "S"), data = dat)
M1$logi.S <- glmer(reformulate(fm$M1[3], response = "H.N"), data = dat, family = binomial)
```

```{r qqplot-families-M1}
diag.M1 <- get.diagnostic.data(M1)
arg <- list(main = "Normality of residuals", xlab = "normal quantiles", ylab = "empirical quantiles", pch = "+")
qqmath(~ Residual | Family, data = diag.M1, scales = list(y = list(relation = "free")),
       xlab = arg$xlab, ylab = arg$ylab, main = arg$main, pch = arg$pch, abline = c(0, 1))
```

```{r scedasticity-families-M1}
arg <- list(main = "Error variance", xlab = "fitted value", ylab = expression(sqrt(residual)), pch = "+")
xyplot(sqrt(abs(Residual)) ~ Fitted.value | Family, data = diag.M1, scales = list(relation = "free"),
       xlab = arg$xlab, ylab = arg$ylab, main = arg$main, pch = arg$pch)
```
