---
layout: default
title: "Trellis Display of Data"
tag: [ multivariate ]
featimg: "S-age-tot-read-count-1.png"
---

Improved graphical overview of data and results pertaining to the recently extended [regression analysis]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-06-17-extending-regression-analysis %}) and [ANOVA]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-06-22-extending-anova %})

## Prepare data

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(lattice)
library(latticeExtra)
library(ggplot2)
library(GGally)
source("../../src/import-data.R")
source("../../src/fit-glms.R")
opts_chunk$set(dpi = 144)
opts_chunk$set(out.width = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
```

```{r cache=TRUE, echo=FALSE}
E <- get.predictors()
gene.ids <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(gene.ids) <- gene.ids
Y.long <- reshape.Y2long(gene.ids)
```

## Dependence of $$S$$ on certain variables

### Dependence on gene, age, and institution

Implementation of the same plot both with the `lattice` and the `ggplot2` package.

```{r S-age-smooth, dev=c("png", "pdf"), warning=FALSE}
P <- list()
# lattice implementation
P$s.age.inst$lattice <-
    xyplot(S ~ Age | Gene, data = Y.long,
           subset = Gene %in% gene.ids,
           groups = Institution,
           panel = function(x, y, ...) {
               panel.xyplot(x, y, pch = 21, cex = 0.3, ...)
               panel.smoother(x, y, col = "black", lwd = 2, ...)
           },
           auto.key = list(title = "institution", columns = 3),
           par.settings = list(add.text = list(cex = 0.8)),
           ylab = "read count ratio, S",
           xlab = "age",
           aspect = "fill", layout = c(6, 5))
# ggplot2 implementation
g <- ggplot(data = Y.long, aes(x = Age, y = S))
g <- g + geom_point(pch = "o", aes(color = Institution))
g <- g + geom_smooth(method = "loess", color = "black")
g <- g + facet_wrap(~ Gene)
P$s.age.inst$ggplot2 <- g
plot(P$s.age.inst$lattice)
plot(P$s.age.inst$ggplot2)
```

### Gene, age, and gender

```{r S-age-Dx, dev=c("png", "pdf"), warning=FALSE}
P <- list()
# lattice implementation
P$s.age.Dx$lattice <-
    xyplot(S ~ Age | Gene, data = Y.long, groups = Dx,
           subset = Gene %in% gene.ids,
           panel = function(x, y, ...) {
               panel.xyplot(x, y, pch = 21, ...)
               #panel.smoother(x, y, col = "black", lwd = 2, ...)
           },
           par.settings = list(add.text = list(cex = 0.8),
                               superpose.symbol = list(cex = 0.5,
                                                       fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)],
                                                       col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
           auto.key = list(title = "Dx", columns = 2),
           ylab = "read count ratio, S",
           xlab = "age",
           aspect = "fill", layout = c(6, 5))
# ggplot2 implementation
g <- ggplot(data = Y.long, aes(x = Age, y = S))
g <- g + geom_point(pch = "o", aes(color = Dx))
g <- g + geom_smooth(method = "loess", color = "black")
g <- g + facet_wrap(~ Gene)
P$s.age.Dx$ggplot2 <- g
plot(P$s.age.Dx$lattice)
#plot(P$s.age.Dx$ggplot2)
```

```{r S-age-gender, dev=c("png", "pdf"), warning=FALSE, echo=FALSE}
P <- list()
# lattice implementation
P$s.age.gender$lattice <-
    xyplot(S ~ Age | Gene, data = Y.long, groups = Gender,
           subset = Gene %in% gene.ids,
           panel = function(x, y, ...) {
               panel.xyplot(x, y, pch = 21, ...)
               panel.smoother(x, y, col = "black", lwd = 2, ...)
           },
           par.settings = list(add.text = list(cex = 0.8),
                               superpose.symbol = list(cex = 0.5,
                                                       fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)],
                                                       col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
           auto.key = list(title = "gender", columns = 2),
           ylab = "read count ratio, S",
           xlab = "age",
           aspect = "fill", layout = c(6, 5))
# ggplot2 implementation
g <- ggplot(data = Y.long, aes(x = Age, y = S))
g <- g + geom_point(pch = "o", aes(color = Gender))
g <- g + geom_smooth(method = "loess", color = "black")
g <- g + facet_wrap(~ Gene)
P$s.age.gender$ggplot2 <- g
plot(P$s.age.gender$lattice)
#plot(P$s.age.gender$ggplot2)
```

```{r S-age, dev=c("png", "pdf"), warning=FALSE}
P$s.age$lattice <-
    xyplot(S ~ Age | Gene, data = Y.long,
           subset = Gene %in% gene.ids,
           par.settings = list(add.text = list(cex = 0.8),
                               strip.background = list(col = "gray90"),
                               plot.symbol = list(pch = 21, cex = 0.5, col = "black", fill = "gray", alpha = 0.5)),
           auto.key = list(title = "gender", columns = 2),
           panel = function(x, y, ...) {
               panel.xyplot(x, y, pch = 21, cex = 0.3, ...)
               panel.smoother(x, y, col = "plum", lwd = 2, ...)
           },
           ylab = "read count ratio, S",
           xlab = "age",
           aspect = "fill", layout = c(6, 5))
plot(P$s.age$lattice)
```

```{r Q-age, dev=c("png", "pdf"), warning=FALSE}
P$s.age$lattice <-
    xyplot(Q ~ Age | Gene, data = Y.long,
           subset = Gene %in% gene.ids,
           par.settings = list(add.text = list(cex = 0.8),
                               strip.background = list(col = "gray90"),
                               plot.symbol = list(pch = 21, cex = 0.5, col = "black", fill = "gray", alpha = 0.5)),
           auto.key = list(title = "gender", columns = 2),
           panel = function(x, y, ...) {
               panel.xyplot(x, y, pch = 21, cex = 0.3, ...)
               panel.smoother(x, y, col = "plum", lwd = 2, ...)
           },
           ylab = "read count ratio, S",
           xlab = "age",
           aspect = "fill", layout = c(6, 5))
plot(P$s.age$lattice)
```

```{r S-age-b, dev=c("png", "pdf"), warning=FALSE, fig.asp=1.5, fig.width=5}
update(P$s.age$lattice, layout = c(5, 6))
```

```{r R-age-gender, dev=c("png", "pdf"), warning=FALSE, echo=FALSE}
xyplot(R ~ Age | Gene, data = Y.long, groups = Gender,
       subset = Gene %in% gene.ids,
       panel = function(x, y, ...) {
           panel.xyplot(x, y, pch = 21, ...)
           #panel.smoother(x, y, col = "black", lwd = 2, ...)
       },
       par.settings = list(add.text = list(cex = 0.8),
                           superpose.symbol = list(cex = 0.5,
                                                   fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)],
                                                   col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
       auto.key = list(title = "gender", columns = 2),
       ylab = "transformed read count ratio, R",
       xlab = "age",
       aspect = "fill", layout = c(6, 5))
```

```{r Q-age-gender, dev=c("png", "pdf"), warning=FALSE, echo=FALSE}
xyplot(Q ~ Age | Gene, data = Y.long, groups = Gender,
       subset = Gene %in% gene.ids,
       panel = function(x, y, ...) {
           panel.xyplot(x, y, pch = 21, ...)
           panel.smoother(x, y, col = "black", lwd = 2, ...)
       },
       par.settings = list(add.text = list(cex = 0.8),
                           superpose.symbol = list(cex = 0.5,
                                                   fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)],
                                                   col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
       auto.key = list(title = "gender", columns = 2),
       ylab = "transformed read count ratio, Q",
       xlab = "age",
       aspect = "fill", layout = c(6, 5))
```

### Dependence on gene, age, and total read count $$N$$

```{r S-age-tot-read-count, dev=c("png", "pdf"), warning=FALSE, echo=FALSE}
# easy lattice implementation
P$s.age.n$lattice <-
    levelplot(log10(N) ~ Age + S | Gene, data = Y.long, alpha = 1.0, col = "white",
              subset = Gene %in% gene.ids,
              ylab = "read count ratio, S",
              xlab = "age, years",
              panel = panel.levelplot.points, layout = c(6, 5),
              colorkey = list(space = "top"), main = expression(paste(log[10], "(total read count)")),
              cex = 0.7,
              par.settings = list(add.text = list(cex = 0.8),
                                  par.main.text = list(cex = 1),
                                  regions = list(col = colorRampPalette(c("blue", "red")))))
#
# alternative lattice implementation without color scale, see Fig 5.5 in 'Lattice: Multivariate Data Visualization with R', http://lmdvr.r-forge.r-project.org
#xyplot(S ~ Age | Gene, data = Y.long, alpha = 0.5, col = "white",
#       groups = colorRampPalette(c("blue", "red"))(100)[ with(Y.long, cut(log10(N), 100, labels = FALSE)) ],
#       panel = function(x, y, subscripts, groups, ...) {
#           fill <- groups[subscripts]
#           panel.xyplot(x, y, pch = 21, fill = fill, ...)
#       })
#
# ggplot2 implementation
g <- ggplot(Y.long, aes(x = Age, y = S, color = log10(N)))
g <- g + geom_point(alpha = I(0.2))
g <- g + scale_color_gradient(low = "blue", high = "red")
g <- g + facet_wrap(~ Gene)
P$s.age.n$ggplot2 <- g
plot(P$s.age.n$lattice)
plot(P$s.age.n$ggplot2)
```

## Associations between explanatory variables

### Deterministic association: RIN and RIN2

```{r rin-rin2, dev=c("png", "pdf"), fig.width=3, out.width = 300, warning=FALSE, echo=FALSE, dpi=72}
xyplot(RIN2 ~ RIN, data = E[sort.int(E$RIN, index.return = TRUE)$ix, ],
       type = "l", grid = TRUE, aspect = 1)
qplot(RIN, RIN2, data = E[sort.int(E$RIN, index.return = TRUE)$ix, ],
      geom = "line", asp = 1)
```

### Stochastic (statistical) associations

Both "scatter plot matrices" show the same set of pairwise associations (top: `lattice`, bottom: `ggplot2` and `GGally` packages).

```{r evar-scatterplot-matrix, dev=c("png", "pdf"), fig.asp=1, warning=FALSE, echo=FALSE, message=FALSE}
P$splom$lattice <-
    splom(E[ e.vars[c(1:6, 10)]], pch = "o", alpha = 0.3, xlab = "", pscales=0,
          panel = function(x, y, ...)
              panel.splom(x, y, ..., jitter.x = ! is.numeric(x), jitter.y = ! is.numeric(y))
          )
#P$splom$ggplot2 <- ggpairs(E[ e.vars[c(1:6, 10)]])
plot(P$splom$lattice)
#plot(P$splom$ggplot2)
```

```{r evar-scatterplot-matrix-gg, dev=c("png", "pdf"), fig.asp=1, warning=FALSE, echo=FALSE, message=FALSE}
ggpairs(E[ e.vars[c(1:6, 10)]], showStrips = TRUE)
```

```{r evar-scatterplot-matrix-simple, dev=c("png", "pdf"), fig.asp=1, warning=FALSE, echo=FALSE, message=FALSE}
P$splom$lattice <-
    splom(E[ e.vars[c(5, 1, 6)]], pch = "o", alpha = 0.5, xlab = "",
          panel = function(x, y, ...)
              panel.splom(x, y, ..., jitter.x = ! is.numeric(x), jitter.y = ! is.numeric(y))
          )
plot(P$splom$lattice)
#plot(P$splom$ggplot2)
ggpairs(E[ e.vars[c(5, 1, 6)]])
```
