---
layout: default
tags: [regression, anova ]
featimg: "ranef-gender-gender-gene-m5-1.png"
---

Predict random regression coefficients of the mixed effect model M5 (the best fitting model).  This uses the ranef function of the lme4 R package.  As the documentation says: "\[ranef is a\] generic function to extract the conditional modes of the random effects from a fitted model object. For linear mixed models the conditional modes of the random effects are also the conditional means."

## Preliminaries

```{r echo=FALSE, message=FALSE}
library(lme4)
library(lattice)
opts_chunk$set(dpi = 144)
opts_chunk$set(out.width = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
source("../../src/import-data.R")
source("../../src/fit-glms.R")
source("2017-07-31-mixed-model-coefs.R")
```

Get selected genes (inferred to be imprinted)

```{r cache=TRUE}
gene.ids <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(gene.ids) <- gene.ids
dat <- merge.data(gene.ids = gene.ids)
```

Get model $$M5$$ formula and fit to data.  $$M6$$ is also fitted.

```{r cache=TRUE}
get.formula <- function(model.name = "M5") {
    x <- read.csv(file = "../../results/M-formulas.csv", stringsAsFactors = FALSE)[[model.name]]
    formula(do.call(paste, as.list(x[c(2, 1, 3)])))
}
M5 <- lmer(get.formula("M5"), data = dat)
M6 <- lmer(get.formula("M6"), data = dat)
```

To recapitulate, the linear predictor of $$M5$$ contains the following terms:

```{r}
formula(M5)
```

Recall that the '1's are intercept terms, the rest are slope terms; each kind has its own sets of random coefficients.

## Results

Only the biologically meaningful terms are analyzed here, which represent the random effects of Gene and Gender:Gene.  Let's look at the coefficients for Gene first:

```{r ranef-gene-m5, fig.asp=1.25}
mydotplot(x = get.coef("Gene", M5), layout = c(3, 2))
```

Next, the coefficients for Gender:Gene.  Since this is an interaction of two factors with 2 and 30 levels, respectively, there are $$60 = 2 \times 30$$ levels and the same number of random coefficients for the intercept term:

```{r ranef-gender-gene-m5, fig.asp=1.25}
mydotplot(x = get.coef("Gender:Gene", M5))
```

This figure summarizes the most important results and is meant for the manuscript.  The panel named as $$(1\mid\mathrm{Gender:Gene})$$ presents the difference between the coefficient for Male and that for Female for each gene: $$b_{g\mathrm{Male}}^{(k)} - b_{g\mathrm{Female}}^{(k)}$$, say, where $$g$$ is a given gene and $$k$$ in the superscript identifies the batch of coefficients associated with $$(1\mid\mathrm{Gender:Gene})$$ .

```{r ranef-gender-gender-gene-m5, fig.asp=0.75}
mydotplot(rbind(get.coef(batch = "Gene", model = M5),
                contrast.coef(cf = get.coef("Gender:Gene"), e.var = "Gender")[[1]]),
          layout = c(3, 1))[c(1, 2, 6)]
```

```{r ranef-gender-gender-gene-m5-all-panels, fig.asp=1.5}
mydotplot(rbind(get.coef(batch = "Gene", model = M5),
                contrast.coef(cf = get.coef("Gender:Gene"), e.var = "Gender")[[1]]),
          layout = c(3, 2))
```
