## Import data

```{r}
diff.e.genes <- read.csv("../../data/fromer-2016-nat-neurosci/nn.4399-S5.csv", skip = 1)
our.genes <- list()
# the selected genes that were analyzed using regression
our.genes$selected <- as.character(read.csv("../../data/genes.regression.new")[[1]])
# the genes with significant association to some biological predictor
our.genes$signif <- as.character(read.csv("../../results/signif-gene-effects-either.csv")[[1]])
```

## Results

The following genes were found to be differentially expressed by Fromer et al **AND**

1. either were merely selected for analysis in our study
1. or, furthermore, we found their parental expression bias to be significantly associated to some biological predictor

```{r}
(both <- lapply(our.genes, intersect, levels(diff.e.genes$Gene.Symbol)))
```

The characteristics of the expression change for the selected genes:

```{r}
diff.e.genes[diff.e.genes$Gene.Symbol %in% both$selected, c("Gene.Symbol", "logFC", "p.value")]
```

Further information on the significant gene(s):

```{r}
read.csv("../../results/signif-gene-effects-either-manual-annot.csv", row.names = 1)[both$signif, ]
```

## Conclusion

Out of the `r length(our.genes$signif)` genes that our study found to be significantly associated to some biological predictor in terms of allelic bias, only `r length(both$signif)` of them, `r both$signif`, was/were found to be differentially expressed by Fromer et al.
