### Missing data due to bug

For each gene $g$ of 16026 genes the higher and lower read counts across $I_g\le 579$ individuals $\{H_{ig}, L_{ig} \,|\, i=1,...,I_g\}$ are stored in a `.csv` file.  However, nearly 3 % of those files are empty due to a bug that resulted in multiple headers in [Ifat's html tables][ifat] (see for instance gene AK3) and to the inability of my converter script to deal with such tables.  To see this:

```{r engine="bash"}
DATADIR="../../data/readcount/"
SEDCMD='s/^.*attila attila\s\+\([[:digit:]]\+\).*\/\([^/]\+\)\.csv$/\1,\2/'
OUTFILE="${DATADIR}/fsize-genes.csv"
[ -f $OUTFILE ] || {
    # create header
    echo "file.size,gene.symbol" > $OUTFILE
    # get file size (in bytes) for each gene under 'genes' directory
    find "${DATADIR}/genes" -name '*.csv' | xargs ls -l | sed "$SEDCMD" >> $OUTFILE
}
cat <<EOF
Number of .csv files

all: $(grep --count '^[[:digit:]]\+,' $OUTFILE )
empty: $(grep --count '^0,' $OUTFILE )
EOF
```

```{r echo=FALSE}
library(lattice)
library(latticeExtra)
opts_chunk$set(dpi = 144)
opts_chunk$set(fig.width = 10)
opts_chunk$set(dev=c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
```

### Import read count data

Load data importer functions:
```{r}
source("../../src/import-data.R")
source("2016-07-19-genome-wide-S.R")
```

The following expressions import $S_{ig}$ for all `r 16026 - 442` genes for which the csv file is nonempty.

```{r}
gene.summary <-
    read.csv("../../data/readcount/summary-all-genes.csv")
rownames(gene.summary) <- gene.summary$Symbol
gene.summary$file.size <-
    read.csv("../../data/readcount/fsize-genes.csv", row.names = 2)[rownames(gene.summary), , drop = TRUE]
gene.ids <- with(gene.summary, as.character(Symbol)[ file.size > 0 ])
```

```{r genomwide-data-import, cache=TRUE, warning=FALSE}
Y <- get.readcounts(gene.ids, g.subsets = list(), sel.var = "S")
Y <- data.frame(lapply(Y, getElement, "S"))
```

For more than half of even the genes $g$ with nonempty files the number $I_g$ of observations (the number of individuals/RNA samples with read count data on $g$) is zero.  In what follows, not only these genes are filtered out but also those with less than 10 observations, indicated by the vertical dashed line on the empirical ECDF plot below.

```{r}
min.n.obs <- 10
```
```{r ecdf-n-obs, echo=FALSE, fig.width=7, fig.asp=2/3}
ecdfplot(sapply(Y, function(y) sum(! is.na(y))), type = c("l", "g"),
         panel = function(...) {
             panel.ecdfplot(...)
             panel.abline(v = min.n.obs, lty = 2)
         }, xlab = "number of observations on gene",
         main = paste0("Read count availability (", as.character(ncol(Y)), " genes)"))
```

```{r echo=FALSE}
# filter genes given the minimum number of allowed observations 'min.n.obs'
ok.genes <- names(Y)[sapply(Y, function(y) sum(! is.na(y)) >= min.n.obs)]
# obtain the ECDF of S_ig for all given genes g
ED <- emp.distr.S(Y[ , ok.genes],
                  ss = seq(from = 0.5, to = 1, length.out = 101),
                  with.density = TRUE)
# order genes according to the value of the ECDF at 0.9
gene.order <- order(sapply(ED[[1]], function(f) f(0.9)))
```

```{r echo=FALSE}
ecdf.val.w <- ED$ecdf.val[ , gene.order[1:1000]]
density.w <- ED$density[ , gene.order[1:1000]]
ED.long <- reshape(ecdf.val.w, v.names = "ECDF", varying = names(ecdf.val.w),
                   timevar = "gene", times = factor(names(ecdf.val.w)),
                   idvar = "s", ids = ED$ss, direction = "long")
density.long <- reshape(density.w, v.names = "density", varying = names(density.w),
                   timevar = "gene", times = factor(names(density.w)),
                   idvar = "s", ids = ED$ss, direction = "long")
ED.long$density <- density.long$density
ED.long$gene <- factor(ED.long$gene, levels = names(ecdf.val.w), ordered = TRUE)
rm(list = c("ecdf.val.w", "density.w", "density.long"))
```

```{r echo=FALSE}
# the genes selected by Ifat
gene.ids.ifat <-
    # 8 genes analyzed by Ifat
    c("PEG3", "INPP5F", "SNRPN", "PWAR6", "ZDBF2", "MEG3", "ZNF331", "GRB10",
      # 5 more genes analyzed by AGK 3/2/16
      "PEG10", "SNHG14", "NAP1L5", "KCNQ1OT1", "MEST",
      # 3 more genes present in data files
      "IGF2", "NLRP2", "UBE3A",
      # 'green' novel 1 MB imprinted genes; note that PWAR6 is already included above
      "TMEM261P1", "AL132709.5", "RP11-909M7.3", "SNORD116-20", "RP13-487P22.1", "hsa-mir-335", "PWRN1")
# retain only 'ok.genes' in 'gene.ids.ifat' and order those according to ECDF at 0.9
# sel.g <- ok.genes[gene.order][ sort(match(gene.ids.ifat[gene.ids.ifat %in% ok.genes[gene.order]], ok.genes[gene.order])) ]
```

```{r ecdf-plot, echo=FALSE, fig.width=7}
(ep <- with(ED,
         xyplot(ecdf.val$PEG10 + ecdf.val$ZNF331 + ecdf.val$ZNF286A + ecdf.val$MAPK9 ~ ss,
                scales = list(y = list(text = letters[1:3])),
                xlab = "s", ylab = expression(hat(F)[g](s)),
                main = "Empirical CDF of selected genes g",
                type = c("g", "s"))))
```

```{r ecdf-levelplot, echo=FALSE, fig.width=7}
tp <- standard.theme()
#tp$regions$col <- rev(tp$regions$col)
lattice.options(default.theme = tp)
sc <- list(x = list(at = seq(1, nrow(ED[[2]]), length.out = 6), labels = 5:10 / 10),
           y = list(draw = FALSE))
(lp <-
    levelplot(as.matrix(ED[[2]][rev(gene.order)]),
              xlab = "s", ylab = "gene rank",
              main = expression(paste("Genome-wide distribution of", S[g])),
              aspect = "fill", scales = sc)[rep(1, 1)])
```

```{r rank-by-ecdf, echo=FALSE, fig.width=7}
ecdf.at.some.s(ED = ED, sel.g = ok.genes[gene.order], pch = ".", ylab = "gene rank")
```

```{r ecdf-levelplot-top, echo=FALSE, fig.width=7}
sc <- list(x = list(at = seq(1, nrow(ED[[2]]), length.out = 6), labels = 5:10 / 10),
           y = list(draw = TRUE))
n.top.g <- 50
(lp <-
    levelplot(as.matrix(ED[[2]][rev(gene.order[seq_len(n.top.g)])]),
              xlab = "s", ylab = paste("top", n.top.g, "genes"),
              aspect = "fill", scales = sc)[rep(1, 1)])
```

```{r rank-by-ecdf-top, echo=FALSE, fig.width=7}
ecdf.at.some.s(ED = ED, sel.g = ok.genes[gene.order[seq_len(n.top.g)]], ylab = "gene rank", type = c("g", "p"))
```

The next plot is meant to serve a consistency check with Ifat's corresponding plot, that is [Fig 1 of the previous manuscript][Fig 1].

```{r compare-to-ifats-fig, echo=FALSE, fig.width=7}
ecdf.at.some.s(ED = ED,
               sel.g = ok.genes[gene.order][ sort(match(
                                                        gene.ids.ifat[gene.ids.ifat %in% ok.genes[gene.order]],
                                                        ok.genes[gene.order])) ],
               xlim = c(1, 0), type = c("h", "p"))
```



[ifat]: http://katahdin.mssm.edu/ifat/web/cm/home
[Fig 1]: https://docs.google.com/presentation/d/1YvpA1AJ-zzir1Iw0F25tO9x8gkSAzqaO4fjB7K3zBhE/edit#slide=id.p4
