---
layout: default
featimg: "Q-Dx-RIN-MEST-density-1.png"
---

This analysis presents plots of the conditional distribution of the read count ratio statistic $$S$$ or its quasi-log transformed version $$Q$$ given each a gene a predictor of interest, which is either *Gender* or *Dx*.  For the predictor *Age* see the previous post [trellis display of data]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-06-26-trellis-display-of-data %}).  In the present post a more detailed conditioning is done, using *RIN* and *Institution* for *MEG3* and *MEST*.  These are genes found in a previous analysis (see [permutation test]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-10-03-permutation-test %})) to be significantly associated to *Gender* and *Dx*, respectively.  The plots below show that the distributions greatly overlap between differing levels of gender or disease even for the most significantly associated genes, and that this holds in case of further conditioning on RIN and Institution.

A distinct but related question is also addressed here, namely the correlation between $$S$$ (or $$Q$$) given one gene and that given another gene when the genes are near each other (in the same imprinted gene cluster) or far away (e.g. on a different chromosome).  However, no clear pattern emerges from this part of the analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(lattice)
library(latticeExtra)
source("../../src/import-data.R")
source("../../src/fit-glms.R")
opts_chunk$set(dpi = 144)
opts_chunk$set(out.width = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
```

## Conditioning on a gene and some predictor(s)

Import data on read counts and predictors and cast them in convenient long format.

```{r cache=TRUE, echo=TRUE}
gene.ids <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(gene.ids) <- gene.ids
Y.long <- reshape.Y2long(gene.ids)
```

### Gene and Gender

```{r}
my.update <- function(dp, ...) {
    update(dp, scales = list(x = list(draw = FALSE, relation = "free")),
           par.settings = list(add.text = list(cex = 0.8)),
           xlab = "read count ratio, S",
           ...)
}
```

#### $$S$$ statistic

```{r S-Gender-density, warning=FALSE}
dp <- densityplot(~ S | Gene, groups = Gender, data = Y.long, plot.points = FALSE)[1:30]
update(my.update(dp, auto.key = list(title = "Gender", text = c("Male", "Female"), columns = 2, points = TRUE, lines = FALSE)),
       par.settings = list(superpose.line = list(fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)],
                                                   col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
       scales = list(draw = FALSE, relation = "free"))
```

The same information in box and whisker representation:

```{r S-Gender-bw, warning=FALSE}
dp <- bwplot(Gender ~ S | Gene, data = Y.long)[1:30]
my.update(dp)
```

#### $$Q$$ statistic

```{r Q-Gender-density, warning=FALSE, echo=FALSE}
dp <- densityplot(~ Q | Gene, groups = Gender, data = Y.long, plot.points = FALSE)[1:30]
update(my.update(dp, auto.key = list(title = "Gender", text = c("Male", "Female"), columns = 2, points = TRUE, lines = FALSE)),
       par.settings = list(superpose.line = list(fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)],
                                                   col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
       scales = list(draw = FALSE, relation = "free"), xlab = "Q")
```

```{r Q-Gender-bw, warning=FALSE, echo=FALSE}
dp <- bwplot(Gender ~ Q | Gene, data = Y.long)[1:30]
update(my.update(dp), xlab = "Q")
```

Only with MEG3, the most significantly affected gene by Gender (see a later post), conditioning on various RIN intervals

```{r}
Y.long$RIN.copy <- Y.long$RIN
Y.long$RIN <- equal.count(Y.long$RIN, number = 3)
```

```{r Q-Gender-RIN-MEG3-density, warning=FALSE}
lp <- densityplot(~ Q | RIN * Institution, data = Y.long, groups = Gender, subset = Gene == "MEG3",
                  scales = list(relation = "free"), plot.points = FALSE,
                  par.settings = list(superpose.line = list(fill = trellis.par.get("superpose.symbol")$fill[c(2, 1)], col = trellis.par.get("superpose.symbol")$col[c(2, 1)])),
                  auto.key=TRUE)
useOuterStrips(lp)
```

```{r Q-Gender-RIN-MEG3-bw, warning=FALSE}
lp <- bwplot(Gender ~ Q | RIN * Institution, data = Y.long, subset = Gene == "MEG3", scales = list(x = list(relation = "free")),
            auto.key=TRUE)
useOuterStrips(lp)
```


### Gene and Dx

#### $$S$$ statistic

For the density plots AFF had to be excluded (for some genes there are too few AFF individuals for kernel density estimation)

```{r S-Dx-density, warning=FALSE, echo=FALSE}
dp <- densityplot(~ S | Gene, groups = Dx, data = Y.long, plot.points = FALSE, subset = Y.long$Dx %in% c("Control", "SCZ"))[1:30]
update(my.update(dp), scales = list(draw = FALSE, relation = "free"),
                 auto.key = list(title = "Dx", columns = 3, points = FALSE, lines = TRUE)) 
```

The same information (extended with AFF) in box and whisker representation:

```{r S-Dx-bw, warning=FALSE, echo=FALSE}
dp <- bwplot(Dx ~ S | Gene, data = Y.long)[1:30]
my.update(dp)
```

```{r S-Dx-strip, warning=FALSE, echo=FALSE}
dp <-
    stripplot(S ~ Dx | Gene, groups = Dx, data = Y.long, subset = Y.long$Dx %in% c("Control", "SCZ", "AFF"),
              plot.points = FALSE, panel = function(x, y,...) {
                  if(FALSE) # violin plot doesn't work for several data pockets
                      panel.violin(x, y,..., col = "gray", border = 0)
                  panel.bwplot(x, y,..., pch = "|")
                  panel.stripplot(x, y,..., jitter.data = TRUE, factor = 4, pch = 21, cex = 0.3)
              })[1:30]
update(my.update(dp), scales = list(x = list(draw = TRUE, relation = "same", rot = 90)),
       xlab = "Dx", ylab = "read count ratio, S", auto.key = list(columns = 3),
       par.settings = list(box.rectangle = list(col = "black", fill = "gray"), box.umbrella = list(col = "black"), plot.symbol = list(cex = 0)))
```


```{r S-Dx-age, warning=FALSE, echo=FALSE}
dp <- xyplot(S ~ Age | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, S",
           panel = function(x, y, ...) {
               panel.xyplot(x, y, ...)
               #panel.smoother(x, y, col = "black", lwd = 2, ...)
           })
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

```{r S-Dx-ancestry-1, warning=FALSE, echo=FALSE}
dp <- xyplot(S ~ Ancestry.1 | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, S")
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

```{r S-Dx-ancestry-2, warning=FALSE, echo=FALSE}
dp <- xyplot(S ~ Ancestry.2 | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, S")
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

```{r S-Dx-ancestry-3, warning=FALSE, echo=FALSE}
dp <- xyplot(S ~ Ancestry.3 | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, S")
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

#### $$Q$$ statistic

```{r Q-Dx-density, warning=FALSE, echo=FALSE}
dp <- densityplot(~ Q | Gene, groups = Dx, data = Y.long, plot.points = FALSE, subset = Y.long$Dx %in% c("Control", "SCZ"))[1:30]
update(my.update(dp, auto.key = list(title = "Dx", columns = 3, points = FALSE, lines = TRUE)), scales = list(draw = FALSE, relation = "free")) 
```

```{r Q-Dx-bw, warning=FALSE, echo=FALSE}
dp <- bwplot(Dx ~ Q | Gene, data = Y.long)[1:30]
update(my.update(dp), xlab = "Q")
```

```{r Q-Dx-strip, warning=FALSE, echo=FALSE}
dp <-
    stripplot(Q ~ Dx | Gene, groups = Dx, data = Y.long, subset = Y.long$Dx %in% c("Control", "SCZ", "AFF"),
              plot.points = FALSE, panel = function(x, y,...) {
                  if(TRUE) # violin plot works in this case
                      panel.violin(x, y,..., col = "gray", border = 0)
                  panel.stripplot(x, y,..., jitter.data = TRUE, factor = 4, pch = 21, cex = 0.3)
              })[1:30]
update(my.update(dp), scales = list(x = list(draw = TRUE, relation = "same", rot = 90)),
       xlab = "Dx", ylab = "read count ratio, Q")
```

```{r Q-Dx-age, warning=FALSE, echo=FALSE}
dp <- xyplot(Q ~ Age | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, Q",
             auto.key = list(columns = 3),
           panel = function(x, y, ...) {
               panel.xyplot(x, y, ...)
               #panel.smoother(x, y, col = "black", lwd = 2, ...)
           })
update(dp, par.settings = list(add.text = list(cex = 0.8)))
```

```{r Q-Dx-ancestry-1, warning=FALSE, echo=FALSE}
dp <- xyplot(Q ~ Ancestry.1 | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, Q")
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

```{r Q-Dx-ancestry-2, warning=FALSE, echo=FALSE}
dp <- xyplot(Q ~ Ancestry.2 | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, Q")
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

```{r Q-Dx-ancestry-3, warning=FALSE, echo=FALSE}
dp <- xyplot(Q ~ Ancestry.3 | Gene, data = Y.long, groups = Dx, pch = 21, cex = 0.3, ylab = "read count ratio, Q")
update(dp, par.settings = list(add.text = list(cex = 0.8)), auto.key = list(columns = 3))
```

Only with MEST, the most significantly affected gene by Dx (see a later post), conditioning on various RIN intervals

```{r Q-Dx-RIN-MEST-density, warning=FALSE, echo=FALSE}
lp <- densityplot(~ Q | RIN * Institution, data = Y.long, groups = Dx, subset = Gene == "MEST",
            scales = list(relation = "free"), plot.points = FALSE,
            auto.key = list(title = "Dx", columns = 3, points = FALSE))
useOuterStrips(lp)
```

```{r Q-Dx-RIN-MEST-bw, warning=FALSE, echo=FALSE}
lp <- bwplot(Dx ~ Q | RIN * Institution, data = Y.long, subset = Gene == "MEST", scales = list(x = list(relation = "free")),
            auto.key=TRUE)
useOuterStrips(lp)
```

## Correlation and gene proximity

```{r}
gs <- read.csv("../../results/gene-clusters.csv", row.names = 1)[-1 * 2:14]
gs.sel <- gs[as.character(read.csv("../../data/genes.regression.new")[ , 1, drop = TRUE]), -1 * c(1, 6) ]
gs.sel$Gene <- factor(rownames(gs.sel), levels = rownames(gs.sel), ordered = TRUE)
gs.sel <- gs.sel[with(gs.sel, order(seg, start)), ]
gs.sel[c("cluster", "start", "gene.type")]
```

Select 3-3 consecutive genes in cluster 27 (chr 14) and 28 (chr 15), respectively.  Produce scatter plot matrix as well as parallel plot first using the $$S$$ statistic and then the $$Q$$ statistic.  These plots show no clear pattern in which correlation would depend on gene proximity.

### $$S$$ statistic

```{r splom-clus-27-28-s}
sel.g <- as.character(gs.sel[gs.sel$seg %in% c(27, 28), "Gene"])[1:6]
splom(m <- sapply(get.readcounts(gene.ids)[sel.g], getElement, "S"), pch = "+", pscales = 0)
parallelplot(m)
```

### $$Q$$ statistic

```{r splom-clus-27-28-q}
splom(m <- sapply(get.readcounts(gene.ids)[sel.g], getElement, "Q"), pch = "+", pscales = 0)
parallelplot(m)
```
