---
layout: default
tags: [ regression, interaction, ]
featimg: "beta-age-cond-wnlm-Q-2-1.png"
---

The present article extends previous inference of regression parameters $$\beta$$ (see [extending ANOVA]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-06-22-extending-anova %})) with a conditional analysis.  Operationally, conditioning takes a subset of all observations (i.e. individuals/RNA preparations) defined by a **condition**.  In the present, rather special, case a condition is defined by a single *level* of a single *factor*-type (i.e. categorical) predictor/explanatory variable.  However, more general conditioning setups could also be explored.  In that case multiple predictors might be selected, of which some might be continuous (i.e. a *covariate*) while others categorical, and for each predictor a possibly compound event (e.g. multiple levels of a factor) might be taken.

The choice of Institution and Gender as selected predictors for conditioning is motivated by the fact that they are categorical with 3 and 2 levels, respectively, and that they are clearly associated with Age (see "scatter plot matrices" in [this article]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-06-26-trellis-display-of-data %})).

## Preparations

Load libraries and my custom functions.

```{r}
library(lattice)
library(latticeExtra)
source("~/projects/monoallelic-brain/src/import-data.R")
source("~/projects/monoallelic-brain/src/fit-glms.R")
```

```{r echo=FALSE}
opts_chunk$set(dpi = 144)
opts_chunk$set(out.height = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
```

Next:

1. import data
1. create subsets (conditioning)
1. fit models (logi.S, wnlm.Q)
1. calculate confidence intervals (CI)

```{r cache=TRUE, warning=FALSE, message=FALSE}
E <- get.predictors() # default arguments
# updated gene set
gene.ids <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(gene.ids) <- gene.ids
Y <- get.readcounts(gene.ids = gene.ids, count.thrs = 0)
# predictors for conditioning
s.p <- c("Institution", "Gender")
# exclude unweighed aggregates UA.8 and UA from fitting
Z <- Y[grep("^UA(.8)?$", names(Y), value = TRUE, invert = TRUE)]
M <- list(logi.S = l.l.do.all.fits.sset(s.p, Z, E, e.vars, "logi.S"),
          wnlm.Q = l.l.do.all.fits.sset(s.p, Z, E, e.vars, "wnlm.Q"))
# regression coefficients: estimates and CI at 99 % level
Betas <- lapply(M, function(l3m) {
    df <- do.call(make.groups,
                  lapply(l3m,
                         function(l2m)
                             do.call(make.groups,
                                     lapply(l2m,
                                            get.estimate.CI, conf.lev = 0.99))))
    names(df)[-(1:5)] <- c("Level", "Cond.Var")
    return(df)
})
```

## Results

```{r echo=FALSE}
L.levels <- c("All", "MSSM", "Penn", "Pitt", "Female", "Male")
Betas$logi.S$Level <- ordered(Betas$logi.S$Level, levels = L.levels)
Betas$wnlm.Q$Level <- ordered(Betas$wnlm.Q$Level, levels = L.levels)
# prepare parameters
sup.polygon <- Rows(trellis.par.get("superpose.polygon"), 1:6)
# customized segment plot
my.segplot <- function(dt, skip = FALSE, ...) {
segplot(ordered(Level, levels = rev(levels(Level))) ~ Lower.CL + Upper.CL | Gene, data = dt,
        subset = Coefficient %in% "Age" & Gene %in% gene.ids,
        scales = list(relation = "free", draw = FALSE),
        level = Level[Coefficient %in% "Age"],
        centers = Estimate[Coefficient %in% "Age" & ! Gene %in% "TMEM261P1"], draw.bands = TRUE,
        panel = function(x, y, ...) {
            panel.segplot(x, y, ...)
            panel.abline(v = 0, col = "red")
        },
        par.settings = list(regions = list(col = sup.polygon$col), strip.background = list(col = "lightgray"), add.text = list(cex = 0.8)),
        col.symbol = "black", pch = "I",
        at = 0:6,
        colorkey = FALSE,
        key = simpleKey(text = levels(dt$Level), space = "right", points = FALSE, rectangles = TRUE),
        xlab = expression(paste("regression coefficient ", beta[age])),
        ylab = "condition",
        skip = skip, layout = c(6, 5),
        ...)[c(1:30)[! skip]]
}
# an alternative implementation without resorting to the latticeExtra package
my.segplot2 <- function(dt, skip = FALSE, ...) {
    with(dt,
         dotplot(ordered(Level, levels = rev(levels(Level))) ~ Estimate | Gene,
                 groups = Level,
                 subset = Coefficient %in% "Age",
                 #subset = Coefficient %in% "Age" & ! Gene %in% "TMEM261P1",
                 scales = list(relation = "free", draw = FALSE),
                 prepanel = function(x, y, subscripts, ...)
                     list(xlim = range(c(Lower.CL[subscripts], Upper.CL[subscripts]), na.rm = TRUE)),
                 panel = function(x, y, groups, subscripts, ...) {
                     fill <- trellis.par.get("superpose.symbol")$fill[ groups[subscripts] ]
                     col <- trellis.par.get("superpose.symbol")$col[ groups[subscripts] ]
                     lsegments(x0 = Lower.CL[subscripts], y0 = y, x1 = Upper.CL[subscripts], y1 = y, col = col, ...)
                     panel.xyplot(x, y, col = col, fill = fill, ...)
                     panel.abline(v = 0, lty = 2)
                 },
                 par.settings =
                     list(strip.background = list(col = "lightgray"), plot.symbol =
                          list(pch = 21), superpose.symbol = list(pch = 21), add.text = list(cex = 0.8)),
                 auto.key = list(space = "right"),
                 xlab = expression(paste("regression coefficient ", beta[age])),
                 ylab = "condition",
                 skip = skip, layout = c(6, 5),
                 ...))[c(1:30)[! skip]]
}
```

In the following plots the effect of 5 conditions (*MSSM*, ..., *Female*) is explored on the inferred $$\beta_\mathrm{age}$$.  *All* denotes the full set of observations (no conditioning), as done before.  Note the following

* two types of regression model were fitted
  * *logi.S*: logistic model fitted to untransformed $$S$$
  * *wnlm.Q*: weighted normal linear model fitted to quasi-log-transformed $$S$$
* the colored bars show confidence intervals (CI) for $$\beta_\mathrm{age}$$
  * some CI are missing (could not be calculated because of too little variance in the response)
* the vertical red line spanning each graph is at $$\beta = 0$$, so when a CI does not "cross" that line we can conclude that the effect of *Age* is significant on the response ($$p < 0.01$$)
* the vertical notch "I" inside the bars marks the least square estimate $$\hat{\beta}_\mathrm{age}$$
* the individual displays (each corresponding to a given gene or aggregate dataset across genes) are scaled differently to emphasize the effects of conditioning.
* the fit for TMEM261P1 did not converge under logi.S so the results are missing; for consistency TMEM261P1 is excluded for wnlm.Q as well

```{r beta-age-cond-wnlm-Q, warning=FALSE, echo=FALSE, message=FALSE}
my.segplot(Betas$wnlm.Q, main = expression(paste("Estimate and 99 % CI for ", beta[age] , " under wnlm.Q")))
```

```{r beta-age-cond-logi-S, warning=FALSE, echo=FALSE, message=FALSE}
skip <- ! sapply(M$logi.S$Marginal$All, `[[`, "converged")[1:30]
my.segplot(Betas$logi.S, skip = skip, main = expression(paste("Estimate and 99 % CI for ", beta[age] , " under logi.S")))
```

The same results plotted with different style

```{r beta-age-cond-wnlm-Q-2, warning=FALSE, echo=FALSE, message=FALSE}
my.segplot2(Betas$wnlm.Q, main = expression(paste("Estimate and 99 % CI for ", beta[age] , " under wnlm.Q")))
```

#### For presentations

```{r beta-age-cond-wnlm-Q-2-present, warning=FALSE, echo=FALSE, message=FALSE, fig.asp=0.7}
update(my.segplot2(Betas$wnlm.Q, main = expression(paste("Estimate and 99 % CI for ", beta[age] , " under wnlm.Q")))[c(5:8, 21:24)], layout = c(4, 2), scales = list(relation = "free", x = list(draw = TRUE)), main = "")
```

```{r beta-age-cond-logi-S-2, warning=FALSE, echo=FALSE, message=FALSE}
my.segplot2(Betas$logi.S, skip = skip, main = expression(paste("Estimate and 99 % CI for ", beta[age] , " under logi.S")))
```

Exclude those genes for which the fit was deemed insufficient.  Requires the file `model-checking.csv`, which was generated in a subsequent analysis.

```{r beta-age-cond-logi-S-2-skip, warning=FALSE, echo=FALSE, message=FALSE}
my.segplot2(Betas$logi.S, skip = ! read.csv("../../results/model-checking.csv")$logi.S.fit.OK, main = expression(paste("Estimate and 99 % CI for ", beta[age] , " under logi.S")))
```

## Conclusion

The main results and their interpretation can be summarized as follows:

* For several genes conditioning qualitatively changes the effect of *Age* on the response in a pattern that is similar under logi.S and wnlm.Q.  This suggests that some assumption(s) that is shared by both models is violated.  The most obvious shared assumption is **linearity**.  In other words, (generalized) linear models fail to capture the complex **dependency structure** that relates predictors to each other and to the response.  Bayesian networks provide the most general framework suitable to deal with such dependency structure.  A less general alternative is using (generalized) linear models extended with interaction terms.
* Under wnlm.Q conditioning has weaker effect than under logi.S.  This suggests that while logi.S is more powerful than wnlm.Q (as expected from a theory-based weighting scheme) it is also less robust for reasons extraneous to nonlinearity.  A logi.S-specific assumption is the sigmoidal **link function** (logistic function), whose inflection point is well beyond several hundred years for all genes and thus far beyond the range of *Age* data.  This need for extrapolation makes logi.S particularly sensitive to **outliers**, which is worsened by the reduction of the number of observations.  That adverse effect is stronger in the case of conditioning on *Penn* or *Pitt* than on *MSSM*, which alone provided $$\approx 55 \%$$ of the observations.  Consistent with this, conditioning on *MSSM* has the smallest effect.  Another potential limitation of the logi.S model is the assumption that the *error distribution* is binomial (see discussion on overdispersion in the [voom paper][voom]).
* Conditioning on the *Penn* or *Pitt* levels of *Institution* has in general larger effect than that on levels of *Gender*.  This is consistent with the stronger association of *Age* to *Institution* than to *Gender*, found previously.

### How to deal with the present limitations?

Suggested alternatives, roughly in increasing model complexity

1. under the present models interpret the results more carefully
  * pro: no further analysis required
    * con: the conclusion that age has effect on allelic imbalance remains weakly established
1. linear modeling with voom/limma
  * pro: (1) allows fitting normal linear models on suitably transformed $$S$$ with the advantage of more power than wnlm.Q and (expectedly) more robustness to nonlinearity than logi.S; also (2), voom/limma uses a more detailed form of data, at the level of SNPs instead of genes, potentially using available information more efficiently; (3) mature `R` packages are available
    * con: substantial additional analysis, including the recovery of the full (i.e. SNP-based) read count data
1. extend present models with interaction terms such as that between *Age* and *Institution*
  * pro: can be performed on the readily available gene-based read count data
    * con: including too few interactions may not suffice correcting for nonlinearity whereas even a few interaction terms greatly increase the number of parameters, resulting in overfitting, therefore this approach might need to be combined with formal model selection implemented in for instance the `stats` package of `R` (see `add1` and `step` functions) to find the best balance between model fit and parsimony
1. use Bayesian networks (BN)
  * pro: even the most complex dependency structure can be captured by a BN
    * con: many details of statistics and software implementation must be worked out

[voom]: http://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29
