---
layout: default
tags: [ reproducible-research ]
---

Here all read count data are fetched from Ifat's website using hacks that aim to extract information from HTML tables.

## Preparing data for analysis in `R`

### Read counts from Ifat's website

The source of read count data is the largest HTML table generated dynamically using the `All genes` option in [Ifat's website][summary].  The output is a `csv` file for each gene within a gene-specific subdirectory.  Due to the specifics of Ifat's html tables and the general difficulty of the `html` to `csv` conversion only the "higher" and "lower" read counts could be formatted in a usable way; the underlying **SNP-specific read counts could not be retrieved** in a clean way.  The same holds for the corresponding genotypes.

```{r, engine='bash', eval=TRUE, echo=TRUE}
# directories
PDIR=$HOME/projects/monoallelic-brain # project directory
RCDIR=$PDIR/data/readcount/ # directory for readcounts
OUTDIR=$RCDIR/genes # subdir for gene-wise readcount

# data files
SUMMARYHTML=$RCDIR/summary-all-genes.html # should already exist
SUMMARYCSV=$RCDIR/summary-all-genes.csv # will be created

# script files
SCRIPT1=$PDIR/src/summary-html2csv
SCRIPT2=$PDIR/src/all-genes-readcounts-for-R 

# run scripts
test -f $SUMMARYHTML || exit 1
test -f $SUMMARYCSV || $SCRIPT1 $SUMMARYHTML > $SUMMARYCSV
test -d $OUTDIR || $SCRIPT2 $SUMMARYCSV 2> /dev/null
echo "number of gene-wise read count datasets:"
find $OUTDIR -type d | wc -l
echo "total size:"
du -sh $OUTDIR

```

### Predictors

The data on predictors (explanatory variables, which can be either covariates or factors) was sent to me by Andy as email attached files.  The following code produces two files: `predictors.csv` is the main data table whereas `RNAseq_ID.DLPFC_RNA_ID.csv` is a mapping between RNAseqID (present in read count tables) and DLPFC_RNA_ID (present in the table of predictors).  Putting together those two is much easier in `R` than in shell scripts.

```{r, engine='bash', eval=TRUE, echo=TRUE}
# directories
PDIR=$HOME/projects/monoallelic-brain # project directory
INDIR=$PDIR/data/ifat/age-dependence # input directory
OUTDIR=$PDIR/data/predictors/ # output directory

# input files from Ifat via Andy
INPRED="$INDIR/DLPFC.ensembl.KNOWN_AND_SVA.ADJUST.SAMPLE_COVARIATES.tsv"
INID="$INDIR/samples.csv"

# script files
SCRIPT=$PDIR/src/predictors-for-R

# run scripts
test -d $OUTDIR || $SCRIPT $INPRED $INID $OUTDIR
echo "output files:"
file $OUTDIR/*
```

## Importing into `R`

### Observed predictors and read counts

We will generate a data frame `E` representing matrix $$E$$ such that each column holds the observed values of some predictor (explanatory variable).  During the fitting of some regression model `R` will automatically construct the design matrix $$X$$ from $$E$$ (or a submatrix of $$E$$ if we omit some predictors) by making contrasts for categorical predictors (i.e. factors).

To get as set of responses for regression, we will also generate a list `Y` of data frames representing the set of matrices $$Y = \{ Y_g \mid g\in\mathcal{G} \}$$ so that each matrix $$Y_g$$ corresponds to gene $$g$$ in a selected set $$\mathcal{G}$$.  The columns of $$Y_g$$ are either the observed values $$H_g$$ (or $$L_g$$) of the "higher" (or "lower") read counts for $$g$$, respectively, or the observed values of derived statistics.  These derived statistics are the total read count $$N_g = L_g + H_g$$, the ratio $$S_g = H_g / N_g$$, whose rank-transformation yields $$R_g$$.  For all matrices $$E, Y_{g_1}, Y_{g_2},...$$ the rows correspond to $$n$$ observations, that is $$n$$ RNA samples, each from a different individual.

$$\mathcal{G}$$ is defined by the `gene.ids` below:
```{r eval=FALSE}
             # 8 genes analyzed by Ifat
gene.ids <- c("PEG3", "INPP5F", "SNRPN", "PWAR6", "ZDBF2", "MEG3", "ZNF331", "GRB10",
             # 5 more genes analyzed by AGK 3/2/16
             "PEG10", "SNHG14", "NAP1L5", "KCNQ1OT1", "MEST",
             # 3 more genes present in data files
             "IGF2", "NLRP2", "UBE3A",
             # 'green' novel 1 MB imprinted genes; note that PWAR6 is already included above
             "TMEM261P1", "AL132709.5", "RP11-909M7.3", "SNORD116-20", "RP13-487P22.1", "hsa-mir-335", "PWRN1")
```

### Aggregations

We will redefine $$Y$$ after appending (horizontally concatenating) to it the set of matrices $$\{Y_{a_k\mathcal{G}_l} \mid k=1,2 \;;\; l=1,2,... \}$$, indexed by combinations of an aggregation method $$a_l$$ and a gene subset $$\mathcal{G}_l\subset\mathcal{G}$$.  $$a_1$$ is aggregation of $$\{S_g \mid g\in\mathcal{G}_l\}$$ (or the corresponding $$\{R_g\}$$) by weighted average (`WA`) using total read count weights $$\{N_g\}$$, whereas $$a_2$$ is unweighted average (`UA`).  Ifat's original analysis combined `UA` with the subset given by `gene.ids[1:8]`.  Thus, for observation $$i$$ the we have $$\bar{S}_{i;k\mathcal{G}_j} = \left( \sum_{g\in\mathcal{G}_l} W_{ig} \right)^{-1} \sum_{g\in\mathcal{G}_l} W_{ig} S_{ig}$$, where $$W_{ig}=N_{ig}$$ if $$k=1$$ (`WA`) and $$W_{ig}=1$$ if $$k=2$$ (`UA`).  Rank transformation of $$\bar{S}_{i;k\mathcal{G}_j}$$ yields $$\bar{R}_{i;k\mathcal{G}_j}$$.

### Filtering

Given threshold $$t$$ (`count.thrs`) and gene $$g$$, an observation $$i$$ is filtered out whenever $$N_{ig}\le t$$.  For weighted average over a subset $$\mathcal{G}_l$$ the filtering rule is $$\sum_{g\in\mathcal{G}_l} N_{ig}\le t$$; consequently for observation $$i$$ the statistic $$S_{ig}$$ may be filtered out from some individual genes $$g\in\mathcal{G}_l$$ but may still contribute to the weighted average $$\bar{S}_{i;k\mathcal{G}_j}$$.  This is not the case for unweighted averages, which are derived from gene-wise statistics only *after* those have been filtered.

### Results

Using a new `R` implementation of data import and processing:
```{r}
source("~/projects/monoallelic-brain/src/import-data.R")
```

```{r cache=TRUE}
# default arguments given explictely to both function calls
E <- get.predictors(f.predictors = "~/projects/monoallelic-brain/data/predictors/predictors.csv",
                      f.rna.ids = "~/projects/monoallelic-brain/data/predictors/RNAseq_ID.DLPFC_RNA_ID.csv")
Y <- get.readcounts(gene.ids = gene.ids,
                           data.dir = "~/projects/monoallelic-brain/data/readcount/genes",
                           count.thrs = 50,
                           sel.obs = row.names(get.predictors()),
                           g.subsets = list(A.8 = gene.ids[1:8], A = gene.ids))
```

The warning (if any) about the arguments to `max` may arise when all observations are filtered out for some gene like for TMEM261P1:
```{r}
sapply(Y, function(y) sum(! is.na(y)))
```

[summary]: http://katahdin.mssm.edu/ifat/web/cm/get_pop_freq.pl
