```{r echo=FALSE}
library(lme4)
library(lattice)
#library(latticeExtra)
opts_chunk$set(dpi = 144)
opts_chunk$set(out.width = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
source("../../src/import-data.R")
source("../../src/fit-glms.R")
```

Selected genes (inferred to be imprinted)

```{r}
gene.ids <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(gene.ids) <- gene.ids
```

Prepare data frame including all `r length(gene.ids)` selected genes

```{r cache=FALSE}
dat <- merge.data(gene.ids = gene.ids)
```

## Model selection

The strategy is to start from a model $M1$ that includes the strongest predictor terms based on previous analysis using separately modeling genes, then confirm the significant effect of these terms.  After this, single---mainly technical---terms are added separately (not yet accumulatively) to $M1$ to asses the effect of these.  The terms that improve the fit according to some criterion will be jointly added to $M1$ resulting in $M2$.  A similar procedure involving two sets of biological terms takes model selection further resulting in $M3$ and $M4$.

### Starting with model $M1$

```{r cache=FALSE}
fo <- Q ~ scale(RIN) + (1 | RNA_batch) + (1 | Institution) + (1 | Institution:Individual) + (scale(Age) | Gene)
(M1 <- lmer(fo, data = dat))
```

```{r cache=TRUE, message=FALSE, warning=TRUE, echo=TRUE}
av <- list()
av$RIN <- anova(update(M1, . ~ . - scale(RIN)), M1)
av$RNA_batch <- anova(update(M1, . ~ . - (1 | RNA_batch)), M1)
av$Institution <- anova(update(M1, . ~ . - (1 | Institution)), M1)
av$Individual <- anova(update(M1, . ~ . - (1 | Institution:Individual)), M1)
av$Age.Gene <- anova(update(M1, . ~ . - (scale(Age) | Gene) + (1 | Gene)), M1)
av$Gene <- anova(update(M1, . ~ . - (scale(Age) | Gene)), update(M1, . ~ . - (scale(Age) | Gene) + (1 | Gene)))
```

```{r}
summarize.anova <- function(av)
    do.call(rbind, lapply(av, function(x)
                          data.frame(Delta.AIC = diff(x$AIC),
                                     Delta.BIC = diff(x$BIC),
                                     Chisq = x$Chisq[2],
                                     df = x[["Chi Df"]][2],
                                     p.Chi = x[["Pr(>Chisq)"]][2])))
summarize.anova(av)
```

#### Notable results

* every term improves model fit by all three criteria (AIC, BIC, $\chi^2$ test)
* in particular, *Age* conditioned on *Gene* leads to the 3rd largest improvement (behind *Gene* and *Individual*)
    * however, this result needs to be confirmed in the context of a more general model 

### Extending $M1$ to $M2$ with technical terms

```{r cache=TRUE, message=FALSE, warning=TRUE, echo=TRUE}
av.1 <- list()
# interactions between random-effect factors
av.1$Gene.Instit <- anova(M1, update(M1, . ~ . + (1 | Gene:Institution)))
av.1$Gene.RNA_batch <- anova(M1, update(M1, . ~ . + (1 | Gene:RNA_batch)))
av.1$RNA_batch.Instit <- anova(M1, update(M1, . ~ . + (1 | RNA_batch:Institution)))
# interactions between random-effect covariates and factors
# RIN
av.1$RIN.Instit <- anova(M1, update(M1, . ~ . - (1 | Institution) + (scale(RIN) | Institution)))
av.1$RIN.RNA_batch <- anova(M1, update(M1, . ~ . - (1 | RNA_batch) + (scale(RIN) | RNA_batch)))
av.1$RIN.Gene <- anova(M1, update(M1, . ~ . - (1 | Gene) + (scale(RIN) | Gene)))
# PMI
av.1$PMI.Instit <- anova(M1, update(M1, . ~ . - (1 | Institution) + (scale(PMI) | Institution)))
av.1$PMI.RNA_batch <- anova(M1, update(M1, . ~ . - (1 | RNA_batch) + (scale(PMI) | RNA_batch)))
av.1$PMI.Gene <- anova(M1, update(M1, . ~ . - (1 | Gene) + (scale(PMI) | Gene)))
av.1$PMI <- anova(M1, update(M1, . ~ . + scale(PMI)))
# Age revisited
av.1$Age.Instit <- anova(M1, update(M1, . ~ . - (1 | Institution) + (scale(Age) | Institution)))
av.1$Age.RNA_batch <- anova(M1, update(M1, . ~ . - (1 | RNA_batch) + (scale(Age) | RNA_batch)))
av.1$Age <- anova(M1, update(M1, . ~ . + scale(Age)))
```

```{r}
summarize.anova(av.1)
```

#### Notable results

* interaction of *Gene* with certain other predictors leads to large improvement in fit
    * this is not surprising since *Gene* has by far the largest impact among other main effects (see above)
    * but the fact that the improvement is due to accounting for interaction between a biological predictor (*Age*) and some technical predictor (*RIN*, *Insitution*, or *PMI*) is difficult to explain mechanistically
* *Age* shows essentially no interaction with *Institution* or *RNA_batch*
    * this is expected: the effect of *Age* should not depend on e.g. the *Institution*; so the result shows the advantage of joint modeling because previous results with separate, gene-based, modeling suggested strong interaction between *Age* and *Institution*

### Extending $M2$ with biological terms

```{r cache=TRUE}
M1b <- update(M1, . ~ . + (1 | Gene:Institution) - (scale(Age) | Gene))
(M2 <- update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) | Gene)))
```

```{r cache=TRUE, message=FALSE, warning=TRUE, echo=TRUE}
av.2 <- list()
# Ancestry
av.2$Ances1 <- anova(M2, update(M2, . ~ . + scale(Ancestry.1)))
av.2$Ances2 <- anova(M2, update(M2, . ~ . + scale(Ancestry.2)))
av.2$Ances3 <- anova(M2, update(M2, . ~ . + scale(Ancestry.3)))
av.2$Ances4 <- anova(M2, update(M2, . ~ . + scale(Ancestry.4)))
av.2$Ances5 <- anova(M2, update(M2, . ~ . + scale(Ancestry.5)))
av.2$Ances1.Gene <- anova(M2, update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) + scale(Ancestry.1) | Gene)))
av.2$Ances2.Gene <- anova(M2, update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) + scale(Ancestry.2) | Gene)))
av.2$Ances3.Gene <- anova(M2, update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) + scale(Ancestry.3) | Gene)))
av.2$Ances4.Gene <- anova(M2, update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) + scale(Ancestry.4) | Gene)))
av.2$Ances5.Gene <- anova(M2, update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) + scale(Ancestry.5) | Gene)))
```

```{r}
summarize.anova(av.2)
```

#### Notable results

* ancestry components are not equally important: only *Ancestry.1* and *Ancestry.3* appears to matter
* their main effect is weaker than their interaction with *Gene*
    * this is expected: the effect of *Age* should not depend on e.g. the *Institution*; so the result shows the advantage of joint modeling because previous results with separate, gene-based, modeling suggested strong interaction between *Age* and *Institution*

### Model $M3$ and $M4$

Based on the above findings the linear predictor for model $M3$ and $M4$ is defined as below, and $M3$ and $M4$ are fitted:

```{r cache=TRUE}
(M3 <- update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(Ancestry.1) | Gene)))
M4 <- update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(PMI) +
                           scale(Ancestry.1) + scale(Ancestry.3) | Gene) +
             scale(Ancestry.1))
```

While the fitting of the more simple model $M3$ converges that of the more complex $M4$ does not.  Therefore $M3$ will be used for subsequent analysis.

### Adding more biological terms

Now add *Gender* and *Dx* either as fixed effect, a random effect (both as main effect and as an interaction term with *Gene*)

```{r eval=TRUE, cache=TRUE, message=FALSE, warning=TRUE, echo=TRUE}
av.3 <- list()
# Gender
av.3$Gender.fix <- anova(M3, update(M3, . ~ . + Gender))
av.3$Gender.ran <- anova(M3, m <- update(M3, . ~ . + (1 | Gender)))
av.3$Gender.Gene <- anova(m, update(m, . ~ . + (1 | Gender:Gene)))
# Dx
av.3$Dx.fix <- anova(M3, update(M3, . ~ . + Dx))
av.3$Dx.ran <- anova(M3, m <- update(M3, . ~ . + (1 | Dx)))
av.3$Dx.Gene <- anova(m, update(m, . ~ . + (1 | Dx:Gene)))
```

Note that adding *Gender* and *Dx* in the following way is not the proper way because they are factors.  Moreover, the fit does not converge.

```{r eval=TRUE, cache=TRUE, message=FALSE, warning=TRUE, echo=TRUE}
# factor playing the role of slope doesn't make much sense
invisible(update(M1b, . ~ . + (Gender + scale(Age) + scale(RIN) + scale(Ancestry.1) | Gene)))
```

#### Results

```{r}
summarize.anova(av.3)
```

Only *Gender*---but not *Dx*---improves the fit a little and only in interaction with *Gene*

### An important result

```{r cache=TRUE}
av.4 <- anova(update(M1b, . ~ . + (scale(RIN) + scale(Ancestry.1) | Gene)),
              update(M1b, . ~ . + (scale(Age) + scale(RIN) + scale(Ancestry.1) | Gene)))
summarize.anova(list(av.4))
```

### Saving model formulas

Save all model formulas in the `results` directory

```{r}
write.csv(data.frame(lapply(list(M1 = M1, M1b = M1b, M2 = M2, M3 = M3, M4 = M4),
                            function(x) as.character(formula(x)))),
          file = "../../results/M-formulas.csv", row.names = FALSE)
```

