---
layout: default
tags:
featimg: "posterior-pp-vs-pval-wnlm-Q-1.png"
---

This analysis compares our findings on the biological regulation of parental expression bias to those published by Perez et al ([eLife 2015;4:e07860](https://elifesciences.org/content/4/e07860)), who presented and applied the [BRAIM model]({{ site.baseurl }}{% post_url /projects/monoallelic-brain/2016-04-14-braim-model %}) to RNA-seq data from the cerebellum of P8 and P60 hybrid mice.

## Preparation

```{r echo=FALSE, warnings=FALSE}
library(lattice)
opts_chunk$set(dpi = 144)
opts_chunk$set(out.width = "700px")
opts_chunk$set(dev = c("png", "pdf"))
lattice.options(default.args = list(as.table = TRUE))
lattice.options(default.theme = "standard.theme")
```

Load functions

```{r}
source("~/projects/monoallelic-brain/src/import-data.R")
source("~/projects/monoallelic-brain/src/fit-glms.R")
source("~/projects/monoallelic-brain/src/graphics.R")
```

### Orthology: matching human and mouse genes

Import "associated names" (symbols) for the selected human genes:

```{r cache=TRUE, warning=FALSE}
h.g.names <- unlist(read.csv("../../data/genes.regression.new", as.is = TRUE))
names(h.g.names) <- h.g.names
```

Import results by Perez et al provided in the first sheet of [Supplementary file 1](https://elifesciences.org/content/4/e07860/supp-material1).

```{r}
perez <- read.csv("../../data/elife-07860/elife-07860-supp1-v2.csv", as.is = TRUE)
```


#### Using AnnotationDbi

```{r message=FALSE, warnings=FALSE, eval=TRUE}
library(AnnotationDbi)
library("org.Hs.eg.db")
library("org.Mm.eg.db")
library("hom.Hs.inp.db")
```

An initial attempt with the `AnnotationDbi::inpIDMapper` function retrieved the id of only 3 mouse orthologs for the 30 selected genes.  As a result of this frustrating result the following conversion hack was done.  Mouse gene names were "humanified" by all-capitalization, i.e. using the rule: Mest (mouse) -> MEST (human).  The detailed operations follow:

Convert for the 30 selected human genes the gene names (stored in `h.g.names`) into Ensemble gene ids.  Then print the name of those genes for which the conversion failed.

```{r, eval=TRUE}
human.eg.ids <- unlist(as.list(org.Hs.egSYMBOL2EG[mappedkeys(org.Hs.egSYMBOL2EG)])[h.g.names])
# the conversion failed for these genes:
names(h.g.names[! h.g.names %in% names(human.eg.ids)])
```

Import all mouse gene ids and symbols, do the conversion hack and check which human genes could not be converted:

```{r, eval=TRUE}
# get all mouse genes
mouse.all <- as.list(org.Mm.egSYMBOL2EG[mappedkeys(org.Mm.egSYMBOL2EG)])
# "humanify" mouse gene names by all-capitalization, i.e. using the rule Mest (mouse) -> MEST (human)
names(mouse.all) <- toupper(names(mouse.all))
# convert mouse gene names (i.e. symbols) to upper case and check for non-matching human gene symbols
names(h.g.names[! h.g.names %in% toupper(as.character(perez$gene_name))])
```

This result shows that `r length(h.g.names[! h.g.names %in% toupper(as.character(perez$gene_name))])` human genes could not be matched to any mouse ortholog.  Below is a more principled and therefore more trusted approach using BioMart.

#### Using BioMart

This has been done with the web-based BioMart tool of Ensemble, which produced the following file

```{r}
h2m <- read.csv("../../data/human-mouse-orthology.csv", as.is = TRUE)
str(h2m)
# remove rows without mouse ortholog; also remove the column "Mouse.orthology.confidence..0.low..1.high."
h2m <- h2m[h2m$Mouse.Ensembl.Gene.ID != "", -3]
nrow(h2m) == length(unique(h2m$Mouse.associated.gene.name))
rownames(h2m) <- h2m$Mouse.associated.gene.name
h2m[-1 * 1:2]
```

## Results

### The results of Perez et al

Bringing the data table into a long format for plotting:

```{r}
vname.pp <- paste0("posterior_mean.pp_", fc <- c("parent", "sex", "cross", "age"), "_effect")
vname.beta <- paste0("posterior_mean.", fc,  "_effect")
perez.l <- # long format
    reshape(perez[c("gene_name", vname.pp, vname.beta)], # extract variables of interest
            v.names = c("P", "beta"),
            direction = "long", varying = list(vname.pp, vname.beta), timevar = "predictor", times = fc, idvar = "id1")
```

Plotted are posterior means of two parameters of the BRAIM model:

1. $$\beta_{jg}$$, the regression coefficient mediating the effect of predictor $$j$$ for the parental bias of gene $$g$$; this is called `posterior_mean.age_effect`, `posterior_mean.sex_effect`,..., in the first sheet of [Supplementary file 1](https://elifesciences.org/content/4/e07860/supp-material1) provided by Perez et al.
1. $$P_{jg}$$, the probability that explanatory variable $$j$$ has large effect on parental bias of $$g$$ (i.e. induces a $$c\times$$ larger variance of $$\beta_{jg}$$); this is called `posterior_mean.pp_age_effect`, `posterior_mean.pp_sex_effect`,...,

```{r perez-p-vs-beta}
xyplot(P ~ beta | predictor, data = perez.l, scales = list(x = list(relation = "free")), pch = ".",
       xlab = expression(paste("E[ ", beta, " | data]")), ylab = "E[ P | data]",
       sub = "Perez et al, using BRAIM model",
       main = "Estimates from mouse cerebellum")
```

This plot shows that the posterior mean of a regression coefficient $$\beta_{jg}$$ need not differ greatly from zero even if $$P_{jg}\approx 1$$ provides a strong evidence that predictor $$j$$ significantly influences gene $$g$$.  This follows from the property of the BRAIM model that $$P_{jg}$$ controls only the variance of $$\beta_{jg}$$ but not its mean.

### Comparison between the two studies

The following expressions focus only two predictors of parental bias: **age** and **gender** because only these two are clearly shared by our human study and the mouse study by Perez et al.

#### Data manuiplations

```{r}
perez.s <- perez[perez$gene_name %in% h2m$Mouse.associated.gene.name, # rows: orthologs of selected human genes
                 c("gene_name", # informative columns
                   "posterior_mean.age_effect", "posterior_mean.pp_age_effect",
                   "posterior_mean.sex_effect", "posterior_mean.pp_sex_effect")]
perez.s <-
    do.call(rbind,
            lapply(unique(perez.s$gene_name),
                   function(n) {
                       df <- perez.s[ perez.s$gene_name == n, ]
                       cbind(data.frame(mouse.gene_name = n),
                             data.frame(t(as.matrix(apply(df[ , -1], 2, mean, na.rm = TRUE)))))
                   }))
perez.s$mouse.gene_name <- as.character(perez.s$mouse.gene_name)
rownames(perez.s) <- h2m[perez.s$mouse.gene_name, "Associated.Gene.Name"]
```

```{r echo=FALSE}
ours <- read.csv("../../results/regr-coefs.csv")
extractor <- function(mtype, coef = "Age") {
    df <- ours[with(ours, Coefficient == coef & Model == mtype), ]
    df$Model <- NULL
    df$Gene <- with(df, factor(Gene, levels = as.character(Gene), ordered = TRUE))
    rownames(df) <- as.character(df$Gene)
    names(df) <- c(colnames(df)[1:2], paste(colnames(df)[3:5], mtype, sep = "."))
    return(df)
}
#
expander <- function(vname, out.vname, df.long = combined, df.short = perez.s) {
    df <- data.frame(rep(NA, nrow(df.long)))
    names(df) <- out.vname
    rownames(df) <- rownames(df.long)
    df[intersect(rownames(df.long), rownames(df.short)), ] <- df.short[[vname]]
    return(df)
}
```

Combine evidence for the effect of age and gender from both our human study and the mouse study by Perez et al.  The two types of evidence are: 

1. p-value for rejecting $$\beta_{jg}=0$$ under the wnlm.Q or logi.S model, calculated either from normal distribution theory or a random permutation test
1. the posterior mean $$E[P_{jg} \mid  \mathrm{data}]$$ 

```{r}
# Age
combined.a <- cbind(extractor("wnlm.Q", coef = "Age"),
                  extractor("logi.S")[-1 * c(1:2)])
combined.a <- cbind(combined.a,
                  expander("posterior_mean.age_effect", "beta.post.mean", df.long = combined.a),
                  expander("posterior_mean.pp_age_effect", "pp.post.mean", df.long = combined.a))
# Gender
combined.g <- cbind(extractor("wnlm.Q", coef = "GenderMale"),
                  extractor("logi.S")[-1 * c(1:2)])
combined.g <- cbind(combined.g,
                  expander("posterior_mean.sex_effect", "beta.post.mean", df.long = combined.g),
                  expander("posterior_mean.pp_sex_effect", "pp.post.mean", df.long = combined.g))
combined <- rbind(combined.a, combined.g)
```

#### The main result

Plotted below is $$E[P_{jg} \mid  \mathrm{data}]$$ from Perez et al against $$\log_{10} p_{jg}$$, the p-value, from our study.

```{r posterior-pp-vs-pval-wnlm-Q}
xyplot(pp.post.mean ~ - log10(p.val.t.dist.wnlm.Q) | Coefficient, data = combined, groups = Gene,
       panel = function(x, y, ..., groups) {
           panel.text(x, y, labels = groups, cex = 0.7, ...)
       },
       strip = strip.custom(factor.levels = c("Age", "Gender")),
       xlim = c(-0.2, 2.5),
       #aspect = 1,
       main = "Evidence for significant effect",
       xlab = expression(paste(log[10], p, ",  present work")),
       ylab = "E[ P | data],  Perez et al"
       )
```

Clearly, the evidence from the mouse and human study for any gene $$g$$ do not match each other.  While there is evidence for some genes from both studies that age has an effect, only our human study detected significant effect of gender.  These results suggests large differences between the studies, which are methodological and/or biological in nature.  In particular:

* the RNA-seq data is modeled differently so the interpretation of the $$E[P_{jg} \mid  \mathrm{data}]$$ (BRAIM) is not identical to $$p_{jg}$$ (wnlm.Q)
* differences in species, age window (relatively young mice vs relatively older people), and brain region
* differing measurement protocols
